### 2022-06-09 article.tex
Update

Here we define paradata as the data that describes the generation of data.
In genetic epidemiology, the data generated is mostly the results 
of an analysis (e.g. predicting a person having a disease),
as done by computer code.
In such context, paradata is usually the scientific paper that
describes what the computer code does.
However, this has the unrealtistic 
assumption that there is a perfect match between the paper and the code.
In this chapter it is argued that the source code should should be supplied,
as this is the true paradata: if the paper and code disagree, it is the
code that has generated the results.
The chapter concludes by some rules how to better code to serve as paradata,
and hence allowing computational research to be truly reproducible.

@@

\begin{table}[h]
  \begin{tabular}{lp{5cm}p{5cm}}
    Term      & Definition                                            & Example                                                         \\
    \hline
    Code      & Body of text to be run by a computer                  & The scripts run by a computational experiment                   \\
    Data      & Individual facts, statistics, or items of information & A SNP that has a significant association                        \\
    Genotype  & The DNA allele at a certain location                  & AA, AC, CG, GT, ...                                             \\
    Paradata  & Data that describes the generation of data            & The code to conclude that a SNP has a significant association   \\
    Phenotype & How an organism looks like in the broadest sense      & The concentration of IL6RA in the blood                         \\
    Metadata  & Data that provides information about other data       & The article that describes an experiment                        \\
    Trait     & A phenotype                                           & The concentration of IL6RA in the blood                         
  \end{tabular}
\end{table}

@@

% \paragraph{Universities produce knowledge}

The currency that universities generate is knowledge, which in turn is distributed
in mostly lectures and academic papers.
Academic papers contain knowledge, either as the
report of measurements or the outcome of an experiment.
These reports are described in English, the current lingua franca of science.

@@

This paper focusses on research fields that uses computation
to do experiments and uses genetic epidemiology as an example.
Genetic epidemiology is a field within biology that, among 
others, measures the spread (hence 'epidemiology') 
of heritable (i.e. genetic) traits, as well as the relation
between having a certain genetic makeup and a certain trait,
where the trait can be any human property, such as weight, height,
the amount of metabolites and having a disease yes/no.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Example study}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

As an example, we'll use a pseudorandomly selected paper
from \cite{ahsan2017relative}. The data used by that paper is
from a population study called the Northern Swedish Population
Health Study (NSPHS) that started in 2010 \cite{igl2010northern}. 
The approximately 1000 participants were initially mostly surveyed
about lifestyle \cite{igl2010northern} and follow-up studies
provided the type of data relevant for this paper, 
which are (1) the genetic data \cite{johansson2013identification},
(2) the concentration of certain proteins in the 
blood \cite{enroth2014strong,enroth2015effect}.

@@

The first type of data, the (final form of the) genetic data,
consists out of single nucleotide polymorphisms (SNPs).
SNPs consist out of a name, a position and a nucleotide. 
DNA (organized into chromosomes and present in every (nucleated) cell, 
see figure \ref{fig:eukakyote_dna}) consists out of many nucleotides, of which there are four types,
called adenosine, cytosine, guanine and thyrosine, all commonly abreviated
as A, C, G and T respectively.
One SNP example is \verb|rs12133641|, which is a SNP located at position 
154,428,283,
where some people have a certain nucleotide. In this case, 
67 percent of the people within this study have an A,
and 33 percent have a G (also from \cite{ahsan2017relative}, Table S3).

@@

The second type of data are concentrations of certain proteins in the 
blood. DNA contains the code for building proteins (see Figure 
\ref{fig:dna_to_protein}), 
as well as the rate
at which a protein is created. Some proteins end up in the blood and
their presence can be used to assess the health of an individual.
IL6RA is a protein [TODO: more info]

The field of genetic epidemiology looks -among others- for
correlations between genetic data and biological traits.
For example, figure \ref{fig:ahsan2017relative_table_2_sub} (from \cite{ahsan2017relative}) shows that
SNP \verb|rs12133641| is highly correlated (p value is 3.0 * 10e-73, 
961 individuals) with protein IL6RA. What this results does
not yet teach us, is how this correlation works, yet
figure \ref{fig:ahsan2017relative_s6} shows us the direction of the association:
the X axis shows the possible genetic make-ups (aka the genotype) of the individuals,
where 0 denotes the individuals with genotype
AA (the individual inherited one adenosine 
from his/here mother and one adenosine from his/her father), 
1.0 denotes AG (one A is inherited from one parent, 
where the G is inherited from the other parent) and 2.0 denotes GG.
From Figure \ref{fig:ahsan2017relative_s6} we can conclude that, on average,
the more guanines are inherited at that SNPs location,
the higher concentration of IL6RA can be found in a human's blood.

The amount of variance that can be explained by an association (i.e.
the R squared value) is rarely 100 percent, which means that a trait (in
this case, IL6RA concentrations) cannot be perfectly explained
from the genotype alone. As we can see in figure \ref{fig:ahsan2017relative_s6}, 
43 percent of the variance can be attributed to an individuals'
genotype and additional factors, 
such as the effect
of the environment (e.g. geographic location, time of day), 
lifestyle (e.g. smoking yes/no) or having a disease (e.g. diabetes) 
are needed to explain the variation between individuals better.

@@

@@

To \emph{obtain} the genetic data, yes, there may be fieldwork and/or lab-work
involved. For example, a researcher needs to go into the field to take
samples (e.g. blood) of humans or other species. To analyse these samples,
a researcher needs a lab. Depending on the technique being used on
how to analyse the samples, there may be a big bioinformatics step
to aggregate the measurements into useful genetic data. 
Although this paper does not focus on the code run to 
aggragate the raw data into useful genetic data, the same arguments
can be made for that code as in the example: that code is 
the ground truth of how the raw data is collected into useful genetic data.

@@

Programming code is paradata, as it is data that describes how data is collected,
which is the definition of paradata [TODO: REF].
Programming code is data that is usually in the form of text, 
spread over one or more files, that describes the experiment.
The experiment collects the data.
The data collected is the data we call the results of an experiment.

The article is metadata, as it is data that describes other
data, which is the definition of metadata [TODO: REF]:
an article describes the
experiment and hence the programming
code in English. However, it is not the best
candidate to describe how the data is collected,
as it has a loose connection with collecting the data.
Again, if the code and article of an experiment disagree,
it is the code that actually let the data be collected.  
Instead, an article is metadata about a research.

>>>>

Here we define paradata as the data that describes the process of generating data.
In genetic epidemiology, the data generated is mostly the results 
of an analysis (e.g. predicting a person having a disease),
as done by computer code.
In such context, one could argue incorrectly that
it is the scientific paper describes what the computer code does.
However, this has the unrealtistic 
assumption that there is a perfect match between the paper and the code.
In this chapter it is argued that the source code should should be supplied,
as this is the true paradata: if the paper and code disagree, it is the
code that has generated the results.
The chapter concludes by some rules how to better code to serve as paradata,
and hence allowing computational research to be truly reproducible.

@@

\begin{table}[h]
  \begin{tabular}{lp{5cm}p{5cm}}
    Term      & Definition                                            & Example                                                         \\
    \hline
    Code      & Body of text to be run by a computer                  & The scripts run by a computational experiment                   \\
    Data      & Individual facts, statistics, or items of information & A SNP that has a significant association                        \\
    Genotype  & The DNA allele at a certain location                  & AA, AC, CG, GT, ...                                             \\
    Paradata  & Data that describes the process of generating data    & The code to conclude that a SNP has a significant association   \\
    Phenotype & How an organism looks like in the broadest sense      & The concentration of IL6RA in the blood                         \\
    Primary data & Data that is worked with in a research             & The genotype of individuals   \\
    Metadata  & Data that provides information about other data       & The article that describes an experiment                        \\
    Raw data  & Data directly from the real world, that is not yet ready to be primary data & The raw results of getting the genotypes of individuals   \\
    Trait     & A phenotype                                           & The concentration of IL6RA in the blood                         
  \end{tabular}
\end{table}

@@

% \paragraph{Simple example}

Two different researchers write two different papers,
that are both accepted. One researcher does not supply the
computer code that was used to generate the results.
The experiment of the other researcher can be re-done easily
and you can verify that, indeed, the results are the same
as in the paper.
Do you trust the conclusion of their papers equally?

In this paper, paradata is defined as information about how data was generated,
as shown schematically in the left panel of Figure \ref{fig:figure_1}.
Paradata does not have a clear definition yet, with common elements
such as that paradata is usually not the primary data, that it describes
a process, that it describes the way of collecting primary data, or that it
describes the way the primary data is generating.

\begin{figure}[!htbp]
  \centering
  \includegraphics[width=\linewidth]{figure_1.png}
  \caption{
    Left: general relations between data, paradata and metadata.
    Middle and right: examples of these types of data
  }
  \label{fig:figure_1}
\end{figure}

% \paragraph{Simple example of paradata}

A simple example would be the data at eBird (\url{https://ebird.org/}),
where people can submit wher a bird of a certain species was spotted.
The name of the person that spotted that bird is paradata, 
as it is that person that generated the data and determined of
which species the bird was, 
as shown schematically in the middle panel of Figure \ref{fig:figure_1}.

% \paragraph{Relevance of paradata}

Paradata is relevant, as it allows primary data to be re-used.
In the eBird example, it helps assess the trustworthiness of the 
data, as beginner birdwatchers make more mistakes in identifying
the correct bird species, as compared to experts.
By using the name and/or experience of the birder as a covariate,
we can re-use the dataset in other contexts as well.

% \paragraph{This paper is about computer code}

This paper is about paradata in the context of a computational
experiment in general and using an example from genetic epidemiology.
A computational experiment is run by computer code
and usually collects primary data first, then runs an analysis,
which results in the generation of results.
This paper takes the standpoint that it is the computer code 
that describes this generation of results, hence computer code is paradata,
as shown schematically in the right panel of Figure \ref{fig:figure_1}.
Reading the computer code allows one to assess the trustworthiness 
of the experiment, similar to knowing the name or expertise of a bird spotter.
Having the code allows one to re-use it, in the same or other contexts,
providing for reproducible research.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Genetic epidemiology}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

% \paragraph{Introduction}

This paper uses genetic epidemiology as a specific example,
to illustrate that the computer code is paradata,
that this paradata is relevant, as it helps 
to assess the correctness of the results,
where the correctness of the results is relevant for healthcare.

@@

Genetic epidemiology is a field within biology that, among 
others, measures the spread (hence 'epidemiology') 
of heritable (i.e. genetic) traits (i.e. a genotype), 
as well as the relation between having a certain genetic makeup 
and a certain trait (i.e. a phenotype),
where the trait can be any human property, such as weight, height,
the amount of metabolites and having a disease yes/no.
Genetic epidemiologists look for associations between genotype
and phenotype. These associations are subsequently used to 
create so-called gene panels, where the gene causing 
an association is measured specifically, to, among others,
detect people at risk for a phenotype.

@@

The example study followed is a pseudorandomly selected paper
from \cite{ahsan2017relative}. The primary data used by that paper is
from a population study called the Northern Swedish Population
Health Study (NSPHS) that started in 2010 (\cite{igl2010northern}). 
The approximately 1000 participants were initially mostly surveyed
about lifestyle (\cite{igl2010northern}) and follow-up studies
provided the type of data relevant for this paper, 
which are (1) the genotypes (\cite{johansson2013identification}),
(2) the phenotypes, which are protein concentrations, as measured
from blood (\cite{enroth2014strong,enroth2015effect}).

@@

The first type of primary data, the genotypes, 
consists out of single nucleotide polymorphisms (SNPs, pronounced as 'snips').
A SNP has a name and a location on the DNA, at which there is a certain nucleotide.
DNA (organized into chromosomes and present in every (nucleated) cell, 
see figure \ref{fig:eukakyote_dna}) 
consists out of many nucleotides, of which there are four types,
called adenosine, cytosine, guanine and thyrosine, all commonly abreviated
as A, C, G and T respectively.
One SNP example is \verb|rs12133641|, which is a SNP located at position 
154,428,283, where 67 percent of the people within this study have an A,
and 33 percent have a G (also from \cite{ahsan2017relative}, Table S3).

@@

The second type of data, the phenotypes, 
are concentrations of certain proteins as obtained from blood. 
DNA contains the code for building proteins (see Figure 
\ref{fig:dna_to_protein}), 
as well as the rate
at which a protein is created. Some proteins end up in the blood and
their presence can be used to assess the health of an individual.
IL6RA is one such protein.

The field of genetic epidemiology looks -among others- for
correlations between genetic data and biological traits.
For example, figure \ref{fig:ahsan2017relative_table_2_sub} 
(from \cite{ahsan2017relative}) shows that
SNP \verb|rs12133641| is highly correlated (p-value is $3.0^{-73}$, 
for $n$ = 961 individuals) with protein IL6RA. What this results does
not yet teach us, is how this correlation works, yet
figure \ref{fig:ahsan2017relative_s6} shows us the direction of the association:
the X axis shows the possible genetic make-ups (aka the genotype) of the individuals,
where 0 denotes the individuals with genotype
AA (the individual inherited one adenosine 
from his/here mother and one adenosine from his/her father), 
1.0 denotes AG (one A is inherited from one parent, 
where the G is inherited from the other parent) and 2.0 denotes GG.
From Figure \ref{fig:ahsan2017relative_s6} we can conclude that, on average,
the more guanines are inherited at that SNPs location,
the higher concentration of IL6RA can be found in a human's blood.

The amount of variance that can be explained by an association (i.e.
the R squared value) is rarely 100 percent, which means that a trait (in
this case, the concentration of IL6RA) cannot be perfectly explained
from the genotype alone. As we can see in figure \ref{fig:ahsan2017relative_s6}, 
43 percent of the variance can be attributed to an individuals'
genotype and additional factors, 
such as the effect
of the environment (e.g. geographic location, time of day), 
lifestyle (e.g. smoking yes/no) or having a disease (e.g. diabetes) 
are needed to explain the variation between individuals better.

@@

These conclusions may end up in the clinic.
Imagine that the level of IL6RA is associated with a disease,
we can create a gene panel that specifically measures
SNP \verb|rs12133641|. In this simple example, it will be easier
to measure the level of IL6RA in the blood, than using a gene panel.
However, there are associations published for many diseases,
in which multiple SNPs contribute to being more likely 
to develope a disease in the future. Here, the phenotype (having 
a disease in the future) is impossible to detect at the present
and associations found in earlier studies are used to create a gene panel.
As creating a gene panel is costly, those associations better be correct.

@@

To \emph{obtain} the raw genetic data, yes, there may be fieldwork and/or lab-work
involved. For example, a researcher needs to go into the field to take
samples (e.g. blood) of humans or other species. To analyse these samples,
a researcher needs a lab. Depending on the technique being used on
how to analyse the samples, there may be a big bioinformatics step
to aggregate the measurements into useful genetic data. 
Although this paper does not focus on the code run to 
aggragate the raw data into useful genetic data, the same arguments
can be made for that code as in the example: that code is 
the ground truth of how the raw data is collected into useful genetic data.

@@

Programming code is paradata, 
as it is data that describes how data is generated.
Programming code is data that is usually in the form of text, 
spread over one or more files, that describes the experiment.
The experiment collects the data.
The data collected is the data we call the results of an experiment.

The article is metadata, as it is data that describes other data:
an article describes the
experiment and hence the programming
code in English. However, it is not the best
candidate to describe how the data is collected,
as it has a loose connection with collecting the data.
Again, if the code and article of an experiment disagree,
it is the code that actually let the data be collected.  
Instead, an article is metadata about a research.
